---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse

# Load janitor
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("C:/Users/willi/OneDrive/Documents/GitHub/data_journalism_2025_spring/lab_03/data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008) |>
  clean_names()
# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("C:/Users/willi/OneDrive/Documents/GitHub/data_journalism_2025_spring/lab_03/data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |>
  clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?
A1:65

```{r}
pg_schools <- payments |>
  filter(str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC") | str_detect(vendor_name, "PG COUNTY PUBLIC S") | str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE") | str_detect(vendor_name, "PRINCE GEORGE'S COUNTY PUBLIC") | str_detect(vendor_name, "BD OF EDUCATION OF PG"))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:2024 had the largest amount. 2025, the year after that, is the lowest. That contrast is very jarring. 

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(total_pgcps = sum(amount)) |>
  arrange(desc(total_pgcps))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. I would probably try and do away with less unique parts of the names. E.g. in Plano-Coudon LLC, shorten it to Plano-Coudoun, and make any Ballinger variant just Ballinger. A new field could be created with the new, standardized names after filtering for things like LLC etc, and then you could use the distinct() function. 
Examples:
"WHITING TURNER CONTRACTING CO' and 'WHITING TURNER CONTRACTING COMPANY'
'PLANO-COUDON LLC' and 'PLANO-COUDON, LLC'
'BALLINGER' and 'THE BALLINGER CO'

```{r}
payments |>
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |>
  group_by(vendor_name) |>
  summarise(total_projects = sum(amount)) |>
  arrange(desc(total_projects))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: I arranged by both total_grants as well as fiscal_year to get two kinds of visuals for this. Grants for stem cell research from the state of Maryland were extremely high in 2023 and 2009, the latest and earliest years included respectively. Between those years and excepting an especially down year in 2021, the number fluctuated between around 900,000 and 600,000 somewhat randomly. 

```{r}
md_grants_loans |>
  filter(str_detect(description, "STEM CELL") | str_detect(description, "Stem Cell")) |>
  group_by(fiscal_year) |>
  summarise(total_grants = sum(amount)) |>
  arrange(desc(total_grants))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 9,709 results for grantees based on the code below. There's probably an average of 2 or so names for each unique grantee so I think the actual number might be about half that. The sheer sum of WMATA grants up top, especially when noting the repeat entries, is fascinating. Frederick County Public Schools being the 11th individual entry is surprising. It's likely because that system might happen to be a uniform row and other counties are more divided into separate duplicate rows and need cleaning, but I'd be curious to see why Frederick County receives so much more in grants if the data is an honest representation. As far as what I would do to clean the data, I really have no idea because of how many rows the data is dealing with. I would have to be able to mutate rows en mass and in many different ways. I would start with standardizing case using str_to_upper or some other function, but I don't know what could practically finish the job.

```{r}
md_grants_loans |>
  filter(zip_code == 20740 | 20742) |>
  group_by(grantee) |>
  summarise(total_grants = sum(amount)) |>
  arrange(desc(total_grants))
```
